An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-03-17 EconGrowth-R-Functions.Rnw
Now including separator pages in presentation.

<<setup_parent, echo=FALSE, message=FALSE, eval=TRUE>>=
require(knitr)
require(lattice)
require(latticeExtra)
require(ggplot2)
require(car)
require(mosaic)
require(xtable)
require(nlmrt)
require(micEconCES)
require(reshape2) # Provides access to data melting. See http://cran.r-project.org/web/packages/reshape2/reshape2.pdf
# tikz allows use of LaTeX formatting and font in graphs. Allows for a consistent look across the paper.
# See http://r-forge.r-project.org/R/?group_id=440 for instructions on installing tikzDevice.
require(tikzDevice) 
trellis.par.set(theme=theme.mosaic())
options(width=75)
options(show.signif.stars=TRUE)
opts_chunk$set(
  # ********************************************************************************************************
  eval=TRUE,
  # ********************************************************************************************************
  dev='tikz',    #Allows LaTeX code in graphical output. E.g., "$y$" for a variable name in the legend for a graph.
  tidy=FALSE,
  comment=NA,
  #Tells whether to cache output from chunks are saved. Cacheing saves time. 
  #However, references to LaTeX tables DO NOT WORK if cacheing is turned on (i.e., cache=TRUE).
  cache=FALSE,
  warning=FALSE,  #Tells whether to show warnings in the output.
  message=FALSE, #Turns off messages for all chunks. Set TRUE on an individual chunk to see it.
  echo=FALSE     #Tells whether to echo code for all chunks. Set TRUE on an individual chunk to see its code.
)

>>>>

<<setup_parent, echo=FALSE, message=FALSE, eval=TRUE>>=
require(knitr)
require(lattice)
require(latticeExtra)
require(ggplot2)
require(car)
require(mosaic)
require(xtable)
require(nlmrt)
require(micEconCES)
require(reshape2) # Provides access to data melting. See http://cran.r-project.org/web/packages/reshape2/reshape2.pdf
# tikz allows use of LaTeX formatting and font in graphs. Allows for a consistent look across the paper.
# See http://r-forge.r-project.org/R/?group_id=440 for instructions on installing tikzDevice.
require(tikzDevice) 
trellis.par.set(theme=theme.mosaic())
options(width=75)
options(show.signif.stars=TRUE)
opts_chunk$set(
  # ********************************************************************************************************
  eval=FALSE,
  # ********************************************************************************************************
  dev='tikz',    #Allows LaTeX code in graphical output. E.g., "$y$" for a variable name in the legend for a graph.
  tidy=FALSE,
  comment=NA,
  #Tells whether to cache output from chunks are saved. Cacheing saves time. 
  #However, references to LaTeX tables DO NOT WORK if cacheing is turned on (i.e., cache=TRUE).
  cache=FALSE,
  warning=FALSE,  #Tells whether to show warnings in the output.
  message=FALSE, #Turns off messages for all chunks. Set TRUE on an individual chunk to see it.
  echo=FALSE     #Tells whether to echo code for all chunks. Set TRUE on an individual chunk to see its code.
)

### 2013-03-17 EconGrowthSeminar.Rnw
Now including separator pages in presentation.

\pgfdeclareimage[height=1.56cm]{university-logo}{CalvinLogo.png}
\logo{\pgfuseimage{university-logo}}

% Delete this, if you do not want the table of contents to pop up at
% the beginning of each subsection:
\AtBeginSubsection[]
{
  \begin{frame}<beamer>{Outline}
    \tableofcontents[currentsection,currentsubsection]
  \end{frame}
}

% If you wish to uncover everything in a step-wise fashion, uncomment
% the following command:

%\beamerdefaultoverlayspecification{<+->}

@@

@@

\begin{frame}{Energy Cost Share is Low} %----------------------------
  \begin{itemize}
    \item{$\sim$10\%}
    \pause
    \item{Energy unimportant for classical economists}
    \vspace{\baselineskip}
    \item{Energy usually ignored}
    \pause
    \vspace{\baselineskip}
    \item{But \dots}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Observation and Questions} %----------------------------
  \begin{itemize}
    \item{Historical economic development paths (US, Western Europe) are energy intensive.}
    \vspace{\baselineskip}
    \pause
    \item{What happens when energy supplies are constrained and prices rise?}
    \item{Will developing countries be able to expand their economies?}
    \vspace{\baselineskip}
    \pause
    \item{How account for energy in economic models?}
  \end{itemize}
\end{frame}

@@

@@

@@

\begin{frame}{The Question} %----------------------------
  Which economic growth model best describes (statistically) the role of energy in economic growth for different types of economies?
\end{frame}

@@

@@

@@

@@

@@

@@

\begin{frame}{Example Cobb-Douglas Fits}

Heun complete.

\end{frame}

@@

\begin{frame}{Parameter Estimation for Cobb-Douglas With Energy} %----------------------------
  \begin{equation*}
    y = \mathrm{e}^{\lambda (t-t_0)}k^{\alpha}l^{\beta}e^{\gamma}
  \end{equation*}
  \begin{itemize}
    \item{Objective: determine $\lambda$, $\alpha$, $\beta$, and $\gamma$ that best fits data}
    \item{Almost identical to Cobb-Douglas without energy}
    \item{Perform grid search over $\gamma$}
    \item{Re-estimate $\lambda$, $\alpha$, and $\beta$ at each value of $\gamma$}
    \item{Respect sum constraint ($\alpha + \beta + \gamma = 1$)}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Constant Elasticity of Substitution (CES)} %----------------------------
  \begin{equation*} \label{eq:micEconCES_model}
    \begin{split}
    y = & \gamma \mathrm{e}^{\lambda(t-t_0)} \\
        & \cdot \left\{\delta \left[\delta_1 k^{-\rho_1} + (1-\delta_1)l^{-\rho_1} \right]^{\rho/\rho_1} 
                        + (1-\delta) e^{-\rho} \right\}^{-1/\rho}
    \end{split}
  \end{equation*}
  \begin{center}
    \begin{tabular}{|c|l|}\hline
      $k$   & $K$/$K_0$, capital stock\\
      $l$   & $L$/$L_0$, labor\\
      $e$   & $E$/$E_0$, any of $q$, $x$, or $u$\\
      \hline
      $\gamma$   &   fitting coefficient\\
      $\lambda$  &   technological progress\\
      $\delta$   &   split between ($kl$) and ($e$)\\
      $\rho$     &   ``strength'' of ($kl$) vs. ($e$)\\
      $\delta_1$   &   split between ($k$) and ($l$)\\
      $\rho_1$     &   ``strength'' of ($k$) vs. ($l$)\\
      \hline
    \end{tabular}
  \end{center}
\end{frame}

@@

@@

\begin{frame}{Conclusions} %----------------------------
  \begin{itemize}
  \item
    Including energy is unwarranted \\
    \scriptsize
    Little statistical evidence that including energy in production functions improves fit.
    \normalsize
  \item
    Including energy is unnecessary \\
    \scriptsize
    Energy is always \emph{already} embedded in capital stock ($K$) and labor ($L$).
    \normalsize
  \vskip0pt plus.5fill
  \item
    Need a modeling method that captures energy's embeddedness \\
    \scriptsize
    System dynamics modeling
    \normalsize
  \end{itemize}
\end{frame}

\begin{frame}{Next Steps (Summer 2013)} %----------------------------
  \begin{itemize}
    \item{Estimate statistical properties for \emph{all} fitting parameters}
    \item{Complete paper}
  \end{itemize}  
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{References}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}{References} %----------------------------
  \bibliographystyle{plainnat}
  \renewcommand{\thepage}{\arabic{page}}
  \tiny
    \bibliography{EconGrowth.bib}
  \normalsize
\end{frame}

>>>>

\pgfdeclareimage[height=1.56cm]{university-logo}{figure_other/CalvinLogo.png}
\logo{\pgfuseimage{university-logo}}

% Delete this, if you do not want the table of contents to pop up at
% the beginning of each subsection:
% \AtBeginSubsection[]
% {
%   \begin{frame}<beamer>{Outline}
%     \tableofcontents[currentsection,currentsubsection]
%   \end{frame}
% }

% If you wish to uncover everything in a step-wise fashion, uncomment
% the following command: 
%\beamerdefaultoverlayspecification{<+->}

@@

\begin{frame} %----------------------------
  \begin{center} \huge
    INTRODUCTION
  \normalsize \end{center}
\end{frame}

@@

\begin{frame}{Energy Cost Share is Low} %----------------------------
  \begin{itemize}
    \item{$\sim$10\%}
    \vspace{\baselineskip}
    \pause
    \item{Energy (relatively) unimportant for classical economists}
    \item{Usually ignored}
    \vspace{\baselineskip}
    \pause
    \item{But \dots}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Observation and Questions} %----------------------------
  \begin{itemize}
    \item{Historical economic development paths (US, Western Europe) are energy intensive.}
    \vspace{\baselineskip}
    \pause
    \item{What happens when energy supplies are constrained and prices rise?}
    \item{Will developing countries be able to expand their economies?}
    \vspace{\baselineskip}
    \pause
    \item{How should we account for energy in economic models?}
  \end{itemize}
\end{frame}

@@

\begin{frame} %----------------------------
  \begin{center} \Large
    Factors of Production
  \normalsize \end{center}
\end{frame}

@@

\begin{frame} %----------------------------
  \begin{center} \Large
    Economic Growth Models
  \normalsize \end{center}
\end{frame}

@@

\subsection[The Question]{The Question} %+++++++++++++++++++++++++++++++++++++++

\begin{frame} %----------------------------
  \begin{center} \Large
    The Question
  \normalsize \end{center}
  Which economic growth model best describes (statistically) the role of energy in economic growth for different types of economies?
\end{frame}

@@

\begin{frame} %----------------------------
  \begin{center} \huge
    SOURCES OF DATA
  \normalsize \end{center}
\end{frame}

@@

\begin{frame} %----------------------------
  \begin{center} \Large
    Economic Data
  \normalsize \end{center}
\end{frame}

@@

\begin{frame} %----------------------------
  \begin{center} \Large
    Energy Data
  \normalsize \end{center}
\end{frame}

@@

\begin{frame} %----------------------------
  \begin{center} \Large
    Historical Data
  \normalsize \end{center}
\end{frame}

@@

\begin{frame} %----------------------------
  \begin{center} \huge
    ANALYSIS OF GROWTH MODELS
  \normalsize \end{center}
\end{frame}

@@

\begin{frame}{Example Cobb-Douglas Fit} %----------------------------
  Heun complete
\end{frame}

@@

@@

\begin{frame}{Cobb-Douglas with Energy Example Fit} %----------------------------
  Heun complete
\end{frame}

\begin{frame}{Constant Elasticity of Substitution (CES)} %----------------------------
  \begin{equation*} \label{eq:micEconCES_model}
    \begin{split}
    y = & \gamma \mathrm{e}^{\lambda(t-t_0)} \\
        & \cdot \left\{\delta \left[\delta_1 k^{-\rho_1} + (1-\delta_1)l^{-\rho_1} \right]^{\rho/\rho_1} 
                        + (1-\delta) e^{-\rho} \right\}^{-1/\rho}
    \end{split}
  \end{equation*}
  \begin{center}
    \begin{tabular}{|c|l|}\hline
      $k$   & $K$/$K_0$, capital stock\\
      $l$   & $L$/$L_0$, labor\\
      $e$   & $E$/$E_0$, any of $q$, $x$, or $u$\\
      \hline
      $\gamma$   &   fitting coefficient\\
      $\lambda$  &   technological progress\\
      $\delta_1$   &   split between ($k$) and ($l$)\\
      $\rho_1$     &   ``strength'' of ($k$) vs. ($l$)\\
      $\delta$   &   split between ($kl$) and ($e$)\\
      $\rho$     &   ``strength'' of ($kl$) vs. ($e$)\\
      \hline
    \end{tabular}
  \end{center}
\end{frame}

@@

\begin{frame} %----------------------------
  \begin{center} \huge
    RESULTS
  \normalsize \end{center}
\end{frame}

@@

\begin{frame} %----------------------------
  \begin{center} \huge
    SUMMARY
  \normalsize \end{center}
\end{frame}

\begin{frame}{Tentative Conclusions} %----------------------------
  \begin{itemize}
  \item{Including energy is unwarranted \\
        \scriptsize
        Little statistical evidence that including energy in production functions improves stastical properties of fitting parameters.
        \normalsize}
  \item{Including energy is unnecessary \\
        \scriptsize
        Energy is \emph{always already} embedded in capital stock ($K$) and labor ($L$).
        \normalsize}
  \vspace{\baselineskip}
  \item{Need a modeling method that captures energy's embeddedness \\
        \scriptsize
        System dynamics modeling
        \normalsize}
  \end{itemize}
\end{frame}

\begin{frame}{Next Steps (Summer 2013)} %----------------------------
  \begin{itemize}
    \item{Estimate statistical properties for \emph{all} fitting parameters}
    \vspace{\baselineskip}
    \vspace{\baselineskip}
    \item{Complete paper}
  \end{itemize}  
\end{frame}

\begin{frame}{References} %----------------------------
  \bibliographystyle{plainnat}
  \tiny
    \bibliography{EconGrowth.bib}
  \normalsize
\end{frame}
